---
title: "Practical Machine Learning"
author: "Ram"
date: "15/01/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This is the final report for Coursera’s Practical Machine Learning course, as part of the Data Science Specialization track offered by John Hopkins.

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. This is the “classe” variable in the training set. We train 4 models: Decision Tree, Random Forest, Gradient Boosted Trees, Support Vector Machine using k-folds cross validation on the training set. We then predict using a validation set randomly selected from the training csv data to obtain the accuracy and out of sample error rate. Based on those numbers, we decide on the best model, and use it to predict 20 cases using the test csv set.


## Loading Data and Libraries

```{r}
library(gbm)
library(rpart.plot)
library(rpart)
library(knitr)
library(randomForest)
library(e1071)
library(lattice)
library(ggplot2)
library(caret)
library(kernlab)
library(rattle)
library(corrplot)
set.seed(1212)
```

```{r}
traincsv <- read.csv("pml-training.csv")
testcsv <- read.csv("pml-testing.csv")

dim(traincsv)
```

```{r}
dim(testcsv)
```

## Cleaning the Data

Removing unnecessary variables. Starting with N/A variables.
```{r}
traincsv <- traincsv[,colMeans(is.na(traincsv)) < .9] #removing mostly na columns
traincsv <- traincsv[,-c(1:7)] #removing metadata which is irrelevant to the outcome
```
Removing near zero variance variables.
```{r}
nvz <- nearZeroVar(traincsv)
traincsv <- traincsv[,-nvz]
dim(traincsv)
```
Now that we have finished removing the unnecessary variables, we can now split the training set into a validation and sub training set. The testing set “testcsv” will be left alone, and used for the final quiz test cases.
```{r}
inTrain <- createDataPartition(y=traincsv$classe, p=0.7, list=F)
train <- traincsv[inTrain,]
valid <- traincsv[-inTrain,]
```

## Creating and Testing the Models

Here we will test a few popular models including: Decision Trees, Random Forest, Gradient Boosted Trees, and SVM. This is probably more than we will need to test, but just out of curiosity and good practice we will run them for comparison.

Set up control for training to use 3-fold cross validation.
```{r}
control <- trainControl(method="cv", number=3, verboseIter=F)
```

## Decision Tree

Model:
```{r}
mod_trees <- train(classe~., data=train, method="rpart", trControl = control, tuneLength = 5)
fancyRpartPlot(mod_trees$finalModel)
```

Prediction:
```{r}
pred_trees <- predict(mod_trees, valid)
cmtrees <- confusionMatrix(pred_trees, factor(valid$classe))
cmtrees
```

## Random Forest

```{r}
mod_rf <- train(classe~., data=train, method="rf", trControl = control, tuneLength = 5)
pred_rf <- predict(mod_rf, valid)
cmrf <- confusionMatrix(pred_rf, factor(valid$classe))
cmrf
```

## Gradient Boost Trees

```{r}
mod_gbm <- train(classe~., data=train, method="gbm", trControl = control, tuneLength = 5, verbose = F)

pred_gbm <- predict(mod_gbm, valid)
cmgbm <- confusionMatrix(pred_gbm, factor(valid$classe))
cmgbm
```

## Support Vector Machines

```{r}
mod_svm <- train(classe~., data=train, method="svmLinear", trControl = control, tuneLength = 5, verbose = F)

pred_svm <- predict(mod_svm, valid)
cmsvm <- confusionMatrix(pred_svm, factor(valid$classe))
cmsvm
```

## Results

The best model is the Random Forest model, with 0.9957519 accuracy and 0.0042481 out of sample error rate. We find that to be a sufficient enough model to use for our test sets.

## Predictions on Test set

Running our test set to predict the classe (5 levels) outcome for 20 cases with the Random Forest model.
```{r}
pred <- predict(mod_rf, testcsv)
print(pred)
```

## Appendix

correlation matrix of variables in training set
```{r}
corrPlot <- cor(train[, -length(names(train))])
corrplot(corrPlot, method="color")
```
Plotting the models
```{r}
plot(mod_trees)
```

```{r}
plot(mod_rf)
```

```{r}
plot(mod_gbm)
```


